Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-23 analysis/paper/paper.Rmd
added linear regression model for size

One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas. There is many discussion about lithic technology using multivariate analysis on shape variables or coefficient of variations on associated metric data. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points. With considerations of a set of metric data, @Buchanan2018; @Buchanan2019 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting. Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel. Other factor, such as low level of cultural innovation in a small group could also lead to an increase in standardization of point shapes [@Okumura2014]. For ceramics, @Topi2017 found a standardized vessel shape from the Casas Grandes culture in northwest Mexico using coefficient of variations for the positions of semi-landmarks across shape groups. They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity.

Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001]. In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity. Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020]. More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007]

Despite the wide use of landmark approach in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that demonstrated the changes in forms from a more angular to rounded shape over time. Since then, EFA has been applied to different archaeological materials. 
Similar to landmark approach, most applications are associated with human remains and animal bones to examine biological variation, and stone artifacts to explore topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].

@@

>>>>

One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas. It is common to use multivariate analysis on shape variables or coefficient of variations on associated metric data to discuss standardization, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points. With considerations of a set of metric data, @Buchanan2018; @Buchanan2019 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting. Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel. Other factor, such as low level of cultural innovation in a small group could also lead to an increase in standardization of point shapes [@Okumura2014]. For ceramics, @Topi2017 found a standardized vessel shape from the Casas Grandes culture in northwest Mexico using coefficient of variations for the positions of semi-landmarks across shape groups. They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity. Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001]. In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity. Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020]. More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

Despite the wide use of landmark approach in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time. Later, @Saragusti2005 introduced several functions for the further calculation of the specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves. To compare resharpening trajectories in the European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA, with considerations of size using multivariate linear regression. More recent applications to stone artifacts seek to answer questions about topological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of British Acheulean [@Hoggard2019].

@@

```{r size-regression}
# calculate centroid size
jpgs_imported_coo_size <-
  jpgs_imported_coo %>% 
  mutate(size =  coo_centsize(jpgs_imported_coo))

# create pcs dataframe for joining
pc1_df_with_filename <-
  bind_cols(filename = rownames(pc1), pc1) %>% 
  rename(c("pc1" = "x")) %>% 
  mutate(filename = parse_number(filename))

pc2_df_with_filename <-
  bind_cols(filename = rownames(pc2), pc2) %>% 
  rename(c("pc2" = "x")) %>% 
  mutate(filename = parse_number(filename))

# make a dataframe for centroid size and join with pc1
jpgs_imported_coo_size_df <-
  as.data.frame(jpgs_imported_coo_size$fac) %>% 
  mutate(filename = parse_number(as.character(filename))) %>% 
  left_join(pc1_df_with_filename) %>% 
  left_join(pc2_df_with_filename)

# size boxplot by period
ggplot(jpgs_imported_coo_size_df,
       aes(size, fct_rev(phase))) +
  coord_flip() +
  geom_boxplot() +
  labs(x = "phase",
       y = "centroid size")

# build regression model
linearMod <- lm( size ~ pc1, data = jpgs_imported_coo_size_df)

ggplot(jpgs_imported_coo_size_df, 
       aes(size, pc1)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ fct_rev(phase)) +
  labs(x = "centroid size")
```
